---
title: "Jonestown Data Analysis"
author: "Alaine Banninga"
date: "4/13/2020"
output: html_document
---

```{r install-packages}
library(tidyverse)
```

```{r upload-data}
jonestown <- read.csv(file = "data/Jonestown.csv")
```

## Distribution of Jonestown Residents: Age, Gender, and Status
### Univariate Summaries of Age, Gender, and Status
```{r Age-histogram}
# histogram of age variable
ageHist <- ggplot(data = jonestown, aes(x = Age)) +
  geom_histogram(bins = 25) +
  ggtitle("Age of Jonestown Residents") +
  xlab("Age (in years)")
ageHist
```

```{r Gender-barplot}
# relabeling gender variable
jonestown$Gender <- fct_collapse(jonestown$Gender,
                                 "Unknown" = c("","u"),
                                 "Female" = "f",
                                 "Male" = "m")

# bar plot of gender
ggplot(data = jonestown, aes(x = Gender)) +
  geom_bar() +
  ggtitle("Gender of Jonestown Residents") +
  xlab("Gender")
```

```{r Status-barplots}
# bar plot: death status counts (specific dates of death)
ggplot(data = jonestown, aes(x = Status)) +
  geom_bar() +
  ggtitle("Death Status of Jonestown Residents (specific dates)") +
  xlab("Death Status") +
  theme(axis.text.x = element_text(angle = 90))


# Collapsing all residents listed as dead in dataset as "Died"
newStatus <- fct_collapse(jonestown$Status,
                          Died = c("Died","Died Prior to 11/18","Died Prior to 11/19","Died Prior to 11/20","Died Prior to 11/21","Died Prior to 11/22","Died Ryan Party"),
                          Survived = "Survived")

# bar plot: death status counts (dead or alive)
ggplot(data = jonestown, aes(x = newStatus)) +
  geom_bar() +
  ggtitle("Death Status of Jonestown Residents (Dead or Alive)") +
  xlab("Death Status") +
  theme(axis.text.x = element_text(angle = 90))
```


### Place of Birth by Age, Gender, and Status
```{r BirthCountry-vs-Age}
# calculating number of residents per birth country
table(jonestown$BirthCountry)
mean(jonestown$Age, na.rm = TRUE) # mean age of Jonestown residents on 11/18/1978

# bar plot: mean age of jonestown resident by birth country
ageByPOB <- jonestown %>%
  group_by(BirthCountry) %>%
  summarise(mean_age = round(mean(Age, na.rm = TRUE), digits = 0)) %>%
  ggplot(aes(x= reorder(BirthCountry, -mean_age), y=mean_age)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    geom_text(aes(label=mean_age), vjust=-0.3, size=3.5)+
    theme_minimal() +
    ggtitle("Mean Age of Jonestown Residents by their Birth Country") +
    ylab("Mean Age (in years)") +
    xlab("Birth Country")
ageByPOB
```

```{r BirthCountry-vs-Gender}
t <- table(jonestown$BirthCountry, jonestown$Gender)

# Percentages of Males, Females, and Unknown Gender Residents by Country
prop.table(t,1)
```

```{r BirthCountry-vs-Status}
t2 <- table(jonestown$BirthCountry, newStatus)

# Death Status Percentages by Country
prop.table(t2,1)
```



### Gender vs Age
```{r Gender-vs-Age}
# bar plot: mean age of jonestown residents by gender
jonestown %>%
  group_by(Gender) %>%
  summarise(mean_age = round(mean(Age, na.rm = TRUE), digits = 0)) %>%
  ggplot(aes(x=reorder(Gender, -mean_age), y=mean_age)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    geom_text(aes(label=mean_age), vjust=-0.3, size=3.5)+
    theme_minimal() +
    ggtitle("Mean Age of Jonestown Residents by Gender") +
    ylab("Mean Age (in years)") +
    xlab("Gender")
```

```{r Gender-vs-Adults-Minors}
# label blank values as "unknown"
jonestown$AdultMinor <- fct_collapse(jonestown$AdultMinor,
                                     "Unknown" = "")

# bar plot showing number of adults and minors by gender
ggplot(jonestown, aes(x = AdultMinor, fill = Gender)) +
    geom_bar(position = position_dodge()) +
    coord_flip() +
    ggtitle("Distribution of Adults and Minors by Gender") +
    ylab("Count") +
    xlab("Adult/Minor Status on 11/18/1978") +
    labs(fill = "Gender")
```

```{r AdultMinor-Gender-Status}
# bar plot: count of each gender by death status, separated by adult/minor status
ggplot(data=jonestown, aes(x = newStatus, fill = Gender)) +
  geom_bar(stat = "count", position = position_dodge()) +
  facet_grid(AdultMinor ~ .)
```



### Guyana Entry vs Status
```{r LengthofTimeSpentinGuyana-vs-Status}
# num days spent in Guyana total
jonestown$daysSpent <- as.numeric(as.Date("1978-11-18") - jonestown$GuyanaEntry)

# box plots of length of time spent in Guyana by death status
ggplot(data = jonestown, aes(x = newStatus, y = daysSpent)) +
  geom_boxplot() +
  ggtitle("Length of Time Spent in Guyana vs Death Status") +
  xlab("Death Status") +
  ylab("Length of time in Guyana (in days)")
```

```{r Location-on-Nov11-1978-vs-Status}
f <- table(jonestown$Location, newStatus)

# Location on 11/18/1978 vs death status
prop.table(f,1)
# Most residents not in Jonestown, Guyana on 11/18/1978 survived
```



### Jonestown Residence vs Status
```{r JTRes-vs-Status}
# collapsing JTResidence into general types
# e.g. Cottage 1 and Cottage 2 are now "Cottage"
jonestown$JTResidence <- fct_collapse(jonestown$JTResidence,
                                    "Not Known" = c("","Not Known"))
jonestown$genRes <- trimws(gsub('[[:digit:]]+', '', jonestown$JTResidence))
jonestown$genRes <- fct_collapse(jonestown$genRes,
                                 "Cottage" = c("Cottage", "Cottage C"))

# percentage bar plot of jonestown residence type vs death status
ggplot(data = jonestown, aes(x = genRes, fill = newStatus)) +
  geom_bar(position = "fill") +
  ggtitle("Jonestown Residence vs Death Status") +
  xlab("Jonestown Residence Type") +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab("Percentage of Residents Died/Survived") +
  labs(fill = "Death Status")
```

```{r JTRes-Status-Gender}
# bar plots of jonestown residence vs death status, separated by gender
ggplot(data=jonestown, aes(x = genRes, fill = newStatus)) +
  geom_bar(stat = "count", position = position_dodge()) +
  facet_grid(Gender ~ .) +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Jonestown Residence vs Death Status by Gender") +
  xlab("Jonestown Residence Type") +
  ylab("Count of Residents Died/Survived") +
  labs(fill = "Death Status")
```





## Place of Birth and its Relationship to Age
### Univariate Summaries of Age, Birth Country, and Birth State
```{r}
# histogram of age
ageHist
```

```{r birthCountry-hist-freq}
# not very useful, most from USA
ggplot(data = jonestown, aes(x = BirthCountry)) +
  geom_bar(stat = "count")

prop.table(table(jonestown$BirthCountry))
# USA overwhelming majority, next largest proportion is Guyana at 2%
```

```{r birthState-bar-freq}
ggplot(data = jonestown, aes(x = BirthState)) +
  geom_bar(stat = "count") +
  ggtitle("Number of Residents per Birth State") +
  xlab("Birth State")

prop.table(table(jonestown$BirthState))
table(jonestown$BirthState)
# highest proportions from CA, TX; lowest from AZ, DE, MA, NB, OR, RI, WI, WY
```

### Mean Age of Residents by their Jonestown Residence
```{r meanAge-per-resType}
# bar plot of mean age of residents by their Jonestown Residence type
jonestown %>%
  group_by(genRes) %>%
  summarise(mean_age = round(mean(Age, na.rm = TRUE), digits = 0)) %>%
  ggplot(aes(x=reorder(genRes, -mean_age), y=mean_age)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    geom_text(aes(label=mean_age), vjust=-0.3, size=3.5)+
    theme_minimal() +
    ggtitle("Mean Age of Residents per Residence Type") +
    ylab("Mean Age (in years)") +
    xlab("Jonestown Residence Type")
```

### Mean Age of Residents by their Birth Country
```{r age-by-birthCountry}
ageByPOB # bar plot made earlier in analysis
```

### Mean Age of US Residents by their Birth State
```{r age-by-birthState}
# bar plot of mean age of US residents by birth state
ageByState <- jonestown %>%
  group_by(BirthState) %>%
  summarise(mean_age = round(mean(Age, na.rm = TRUE), digits = 0)) %>%
  ggplot(aes(x=reorder(BirthState, -mean_age), y=mean_age)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    geom_text(aes(label=mean_age), vjust=-0.3, size=3.5)+
    theme_minimal() +
    ggtitle("Mean Age of Jonestown Residents by their Birth State") +
    ylab("Mean Age (in years)") +
    xlab("Birth State") +
    theme(axis.text.x = element_text(angle = 90))
ageByState
```

